---
title: "dplyr Tutorial"
author: "Rafael Irizarry, Michael Love, Arrange by Wayne Ho"
date: "2019/7/10"
output: html_document
---

##### **Original Website: <a href='https://genomicsclass.github.io/book/pages/dplyr_tutorial.html'>here</a>**

<h2 id = '0'></h2>

* [Intro](#1)
  
* [Data: mammals sleep](#2)

* [Important dplyr verbs to remember](#3)

* [dplyr verbs in action](#4)
    + [select()](#4.1)
    + [filter()](#4.2)

* [Pipe operator: %>%](#5)

* [Back to dplyr verbs in action](#6)
    + [arrange()](#6.1)
    + [mutate()](#6.2)
    + [summarise()](#6.3)
    + [group_by()](#6.4)
    
<h2 id = '1'></h2>

## **Intro**

<blockquote>

**What is *dplyr*?**
   
>dplyr is a powerful R-package to transform and summarize tabular data with rows and columns.    


**Why is it useful?**
    
>The package contains a set of functions (or “verbs”) that perform common data manipulation operations such as filtering for rows, selecting specific columns, re-ordering rows, adding new columns and summarizing data.In addition, dplyr contains a useful function to perform another common task which is the “split-apply-combine” concept. We will discuss that in a little bit.    
    
**How does it compare to using base functions R?**
    
>If you are familiar with R, you are probably familiar with base R functions such as split(), subset(), apply(), sapply(), lapply(), tapply() and aggregate(). Compared to base functions in R, the functions in dplyr are easier to work with, are more consistent in the syntax and are targeted for data analysis around data frames instead of just vectors.

**How do I get *dplyr*?**
    
>To install *dplyr*

<pre><code>install.packages("dplyr")</code></pre>
    
>To load *dplyr*

<pre><code>library(dplyr)</code></pre>

</blockquote>

[Back to Top](#0)

------------------------------

<h2 id = '2'></h2>

## **Data: mammals sleep**   

<blockquote>
The msleep (mammals sleep) data set contains the sleeptimes and weights for a set of mammals and is available in the dagdata repository on github. This data set contains 83 rows and 11 variables.

Download the msleep data set in CSV format from <a href="https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv">here</a>, and then load into R:
    
```{r warning=FALSE}

library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
filename <- "msleep_ggplot2.csv"
if (!file.exists(filename)) download(url,filename)
msleep <- read.csv("msleep_ggplot2.csv")
head(msleep)

```

The columns (in order) correspond to the following:

column name |	Description
----------- | ------------
name        |	common name
genus       |	taxonomic rank
vore        | 	carnivore, omnivore or herbivore?
order       |	taxonomic rank
conservation|	the conservation status of the mammal
sleep_total |	total amount of sleep, in hours
sleep_rem   |	rem sleep, in hours
sleep_cycle |	length of sleep cycle, in hours
awake     	| amount of time spent awake, in hours
brainwt     |	brain weight in kilograms
bodywt      |	body weight in kilograms

</blockquote>

[Back to Top](#0)

----------------------------------

<h2 id = '3'></h2>
## **Important dplyr verbs to remember**  

<blockquote>

dplyr verbs |	Description
----------- | ------------
select()    |	select columns
filter()    |	filter rows
arrange()   |	re-order or arrange rows
mutate()    |	create new columns
summarise()	|summarise values
group_by()  |	allows for group operations in the “split-apply-combine” concept

</blockquote>

[Back to Top](#0)

-------------------------

<h2 id = '4'></h2>
## **dplyr verbs in action**  

<blockquote>

The two most basic functions are select() and filter() which selects columns and filters rows, respectively.

<h2 id = '4.1'></h2>

**Selecting columns using select()**

<blockquote>
Select a set of columns: the name and the sleep_total columns.

```{r warning=FALSE, message=FALSE}

library(dplyr)
sleepData <- select(msleep, name, sleep_total)
head(sleepData)

```

To select all the columns except a specific column, use the “-“ (subtraction) operator (also known as negative indexing)

```{r}
head(select(msleep, -name))
```

To select a range of columns by name, use the ":" (colon) operator

```{r}
head(select(msleep, name:order))
```

To select all columns that start with the character string “sl”, use the function starts_with()

```{r}
head(select(msleep, starts_with("sl")))
```

Some additional options to select columns based on a specific criteria include

dplyr verbs |	Description
----------- | ------------
ends_with() | Select columns that end with a character string
contains()  | Select columns that contain a character string
matches()   | Select columns that match a regular expression
one_of()    | Select columns names that are from a group of names

</blockquote>

<h2 id = '4.2'></h2>
**Selecting rows using filter()**

<blockquote>

Filter the rows for mammals that sleep a total of more than 16 hours.

```{r}
filter(msleep, sleep_total >= 16)
```

Filter the rows for mammals that sleep a total of more than 16 hours and have a body weight of greater than 1 kilogram.

```{r}
filter(msleep, sleep_total >= 16, bodywt >= 1)
```

Filter the rows for mammals in the Perissodactyla and Primates taxonomic order

```{r}
filter(msleep, order %in% c("Perissodactyla", "Primates"))
```

You can use the boolean operators (e.g. >, <, >=, <=, !=, %in%) to create the logical tests.

</blockquote>

</blockquote>

[Back to Top](#0)

----------------

<h2 id = '5'></h2>
## **Pipe operator: %>%**

<blockquote>

Before we go any futher, let’s introduce the pipe operator: %>%. dplyr imports this operator from another package (magrittr). This operator allows you to pipe the output from one function to the input of another function. Instead of nesting functions (reading from the inside to the outside), the idea of of piping is to read the functions from left to right.
    
Here’s an example you have seen:

```{r}
head(select(msleep, name, sleep_total))
```

Now in this case, we will pipe the msleep data frame to the function that will select two columns (name and sleep_total) and then pipe the new data frame to the function head() which will return the head of the new data frame.

```{r}
msleep %>% 
    select(name, sleep_total) %>% 
    head
```

You will soon see how useful the pipe operator is when we start to combine many functions.

</blockquote>

[Back to Top](#0)

-------------------------

<h2 id = '6'></h2>
## **Back to dplyr verbs in action**

<blockquote>

Now that you know about the pipe operator (%>%), we will use it throughout the rest of this tutorial.

<h2 id = '6.1'></h2>
**Arrange or re-order rows using arrange()**

<blockquote>

To arrange (or re-order) rows by a particular column such as the taxonomic order, list the name of the column you want to arrange the rows by

```{r}
msleep %>% arrange(order) %>% head
```

Now, we will select three columns from msleep, arrange the rows by the taxonomic order and then arrange the rows by sleep_total. Finally show the head of the final data frame

```{r}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, sleep_total) %>% 
    head
```

Same as above, except here we filter the rows for mammals that sleep for 16 or more hours instead of showing the head of the final data frame

```{r}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, sleep_total) %>% 
    filter(sleep_total >= 16)
```

Something slightly more complicated: same as above, except arrange the rows in the sleep_total column in a descending order. For this, use the function desc()

```{r}
msleep %>% 
    select(name, order, sleep_total) %>%
    arrange(order, desc(sleep_total)) %>% 
    filter(sleep_total >= 16)
```

</blockquote>

<h2 id = '6.2'></h2>
**Create new columns using mutate()**

<blockquote>

The mutate() function will add new columns to the data frame. Create a new column called rem_proportion which is the ratio of rem sleep to total amount of sleep.

```{r}
msleep %>% 
    mutate(rem_proportion = sleep_rem / sleep_total) %>%
    head
```

You can many new columns using mutate (separated by commas). Here we add a second column called bodywt_grams which is the bodywt column in grams.

```{r}
msleep %>% 
    mutate(rem_proportion = sleep_rem / sleep_total, 
           bodywt_grams = bodywt * 1000) %>%
    head
```

</blockquote>

<h2 id = '6.3'></h2>
**Create summaries of the data frame using summarise()**

<blockquote>

The summarise() function will create summary statistics for a given column in the data frame such as finding the mean. For example, to compute the average number of hours of sleep, apply the mean() function to the column sleep_total and call the summary value avg_sleep.

```{r}
msleep %>% 
    summarise(avg_sleep = mean(sleep_total))
```

There are many other summary statistics you could consider such sd(), min(), max(), median(), sum(), n() (returns the length of vector), first() (returns first value in vector), last() (returns last value in vector) and n_distinct() (number of distinct values in vector).

```{r}
msleep %>% 
    summarise(avg_sleep = mean(sleep_total), 
              min_sleep = min(sleep_total),
              max_sleep = max(sleep_total),
              total = n())
```

</blockquote>

<h2 id = '6.4'></h2>
**Group operations using group_by()**

<blockquote>

The group_by() verb is an important function in dplyr. As we mentioned before it’s related to concept of “split-apply-combine”. We literally want to split the data frame by some variable (e.g. taxonomic order), apply a function to the individual data frames and then combine the output.

Let’s do that: split the msleep data frame by the taxonomic order, then ask for the same summary statistics as above. We expect a set of summary statistics for each taxonomic order.

```{r}
msleep %>% 
    group_by(order) %>%
    summarise(avg_sleep = mean(sleep_total), 
              min_sleep = min(sleep_total), 
              max_sleep = max(sleep_total),
              total = n())
```

</blockquote>

</blockquote>

[Back to Top](#0)

